tstrait: a quantitative trait simulator for ancestral recombination graphs

### 2024-03-11 paper.tex
Minor changes from Daiki

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
GWAS models~\citep{uffelmann2021}, adapted to the ARG context.
Each trait is associated with one or more causal sites
(positions on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide) associated with an effect size $\beta$.
For each causal site, an effect size is drawn from a distribution
and optionally multiplied by $\left(2p(1-p)\right)^\alpha$,
where $p$ is the frequency of the causal allele and $\alpha$
is a parameter describing the strength of
frequency dependence~\citep{speed2017}.
At a particular causal site, every node in the local tree that inherits
the causal allele at that site is said to have a ``genetic value'' of $\beta$.
% NOTE removing this for space - it keeps this section on one page.
% Because we are reasoning
% about the inheritance of alleles at a site on the genome,
% we can consider just the local tree at that site and
% the network structure of the ARG is not important~\citep{wong2023general}.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales on standard computers.
Supplementary Fig~\ref{fig:time} shows how trait simulation time scales
with number of individuals on human-like coalescent simulations
generated using~\texttt{stdpopsim}~\citep{adrion2020}.
For each ARG, we simulate a trait with 1000 causal sites
and report the mean CPU (Intel(R) Core(TM) i9-11900H)
time over 10 independent replicates.
To emphasise scalability, we also
applied \texttt{tstrait} to the simulations of 1.4 million
French-Canadians discussed in the introduction.
% TODO redo this for 100 causal sites @daiki.
It took 6 seconds to simulate a trait with 100 causal sites.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for an ARG estimated from 1000 Genomes project
data~\citep{kelleher2019} which has YYY samples
and ZZZ variant sites. This took 0.X seconds.
Memory requirements for \texttt{tstrait} are modest: all of the
above experiments were performed on a laptop computer with 16GB of RAM.

@@

\begin{figure*}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) CPU time with increasing number of individuals
based on ARGs simulated using \texttt{stdpopsim} with the [MODEL ID]
demographic mode. For each replicate we simulate
one quantitative trait with 1000 causal sites.
(B) CPU time with increasing number of causal sites on two large ARGs.
\texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure*}

\end{document}

>>>>

\subsection{Model}
Phenotypes are simulated in \texttt{tstrait} following standard
GWAS models~\citep{uffelmann2021}, adapted to the ARG context.
Each trait is associated with one or more causal sites
(positions on the genome), and at each causal site there is a causal allele
(i.e.\ a particular nucleotide) associated with an effect size $\beta$.
For each causal site, an effect size is drawn from a distribution
and optionally multiplied by $\left(2p(1-p)\right)^{\alpha/2}$,
where $p$ is the frequency of the causal allele and $\alpha$
is a parameter describing the strength of
frequency dependence~\citep{speed2012}.
At a particular causal site, every node in the local tree that inherits
the causal allele at that site is said to have a ``genetic value'' of $\beta$.
% NOTE removing this for space - it keeps this section on one page.
% Because we are reasoning
% about the inheritance of alleles at a site on the genome,
% we can consider just the local tree at that site and
% the network structure of the ARG is not important~\citep{wong2023general}.

@@

\subsection{Performance}
\texttt{Tstrait} is very efficient, and can be applied to datasets at the
largest scales on standard computers.
Supplementary Fig~\ref{fig:time} shows how trait simulation time scales
with number of individuals on human-like coalescent simulations
generated using~\texttt{stdpopsim}~\citep{adrion2020}.
For each ARG, we simulate a trait with 1000 causal sites
and report the mean CPU (Intel(R) Core(TM) i9-11900H)
time over 10 independent replicates.
To emphasise scalability, we also
% TODO check that the numbers align here, we may need to be 
% less specific to avoid confusing readers or having to
% explain technical details.
applied \texttt{tstrait} to the simulations of 2.7 million
French-Canadians discussed in the introduction.
% tstrait simulates phenotypes for 2.7 million individuals in the French Canadian dataset, as ancestors are denoted as individuals in the dataset.
% TODO redo this for 100 causal sites @daiki.
It took 80.69 seconds to simulate a trait with 100 causal sites.
% Is this worth the space? We should cite 1KG project also really.
Finally, to demonstrate that \texttt{tstrait} can also be applied
to ARGs inferred from real data, we simulated a trait with 100 causal
sites for an ARG estimated from 1000 Genomes project
data~\citep{kelleher2019} which has 2,504 samples
and 1,685,401 variant sites. This took 5.40 seconds.
Memory requirements for \texttt{tstrait} are modest: all of the
above experiments were performed on a laptop computer with 16GB of RAM.

@@

\begin{figure*}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=-1$. (A) CPU time with increasing number of individuals
based on ARGs simulated using \texttt{stdpopsim} with the HomSap
demographic mode. For each replicate we simulate
one quantitative trait with 1000 causal sites.
(B) CPU time with increasing number of causal sites on two large ARGs.
\texttt{tstrait} is used to simulate quantitative traits from
the simulated French Canadian dataset \citep{anderson2023} and the inferred
tree sequence dataset of the 1000 Genomes Project \citep{kelleher2019} with
varying number of causal sites. The French Canadian dataset is downloaded from
\url{https://zenodo.org/record/6839683}, and the inferred ARG from the 1000
Genomes Project is downloaded from \url{https://zenodo.org/record/3051855}.
Chromosome 9 is selected for both datasets.}\label{fig:time}
\end{figure*}

\begin{figure*}[t]%
\centering
% FIXME get the sizes right here and in the notebook where we're generating
% the figure so we don't have to scale here.
\includegraphics[width=213pt]{figures/time-scaling.pdf}
\caption{\textbf{Time taken to simulate quantitative traits.} Each point
represents the mean time for 10 independent runs under different random seeds with 95\% confidence intervals plotted.
The times reported are the total CPU time required to simulate quantitative
traits on an Intel(R) Core(TM) i9-11900H CPU and 16 GB of RAM. The trait model
is a normal distribution with $\mu=0$, $\sigma^2=1$, $h^2=0.3$, and
$\alpha=0$. \texttt{tstrait} is used to simulate quantitative traits from ARGs simulated using \texttt{stdpopsim} with the HomSap demographic mode. For each replicate we simulate one quantitative trait with 1000 causal sites. We have compared \texttt{tstrait} against the quantitative trait simulation framework that is based on \texttt{tskit} functionalities. The comparison codes are adapted from \citep{martin2017}, and they are modified based on functionalities offered in \texttt{tskit} version 0.5.6 to speed up the computational efficiency.}\label{fig:time}
\end{figure*}
\end{document}
